---
title: Plot Missing Values
author: Masood Sadat
date: '2018-05-22'
slug: plot-missing-values
categories:
  - R tips
tags:
  - R tips
---

In one of [**DataCamp**](http://datacamp.com/) courses that I cannot remember which one, the instructor used `geom_tile` from `ggplot2` package to visualize missing values. Here, I am going to replicate that using `airquality` dataset from base R.

To start, we can use `summary()` and `str()` functions to inspect the data.

```{r message=FALSE, warning=FALSE}
summary(airquality)
```

```{r message=FALSE, warning=FALSE}
str(airquality)
```

From above, I can inspect data type, desriptive statistics, and first few values of each column/variable. The `summary()` also provide if any of the columns have missing values. `glimpse()` from **dplyr** also provide another function to inspect the data.

We found out that `Ozone` and `Solar.R` columns have missing values. 

```{r message=FALSE, warning=FALSE}
library(tidyverse)

airquality %>%
  is.na() %>%
  data.frame() %>%
  mutate(obs = 1:nrow(.)) %>%
  gather(var, is_missing, -obs) %>%
  ggplot(aes(var, obs, fill = is_missing)) + geom_tile()
```







