---
title: Hypothesis testing with R
author: Masood Sadat
date: '2018-06-14'
slug: hypothesis-testing-with-r
categories:
  - r tips
tags:
  - inferential statistics
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(ggplot2)
library(broom)
library(dplyr)
```

### A Meaningful Relationship

The relatinship between weight (`wt`) and mileage (`mpg`) is negative one.

```{r}
# library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
```

Fit a linear model to estimate the linear relationship between weight and mpg. The relationship is negative and it is statistically significant.

```{r}
# library(broom)
# library(dplyr)
tidy(lm(mpg ~ wt, data = mtcars))
obs_slope <- tidy(lm(mpg ~ wt, data = mtcars)) %>% .[2,2]
```

In practice, to do inference, we will need to know the distribution of the slope under the null hypothesis, that is there is no relationship between explanatory and response variables.

> _`pull()` function takes a data frame and returns a selected column as a vector (similar to $)_

```{r}
library(infer)
perm_slopes <- mtcars %>%
  specify(formula = mpg ~ wt) %>% # specify the linear model
  hypothesize(null = "independence") %>% # provide null hypothesis that two variables are independent
  generate(reps = 100, type = "permute") %>% # describe how to generate the sampling distribution
  calculate(stat = "slope") # calculate the statistic of interest
```

Plot the permuted slopes using a histogram and add the observed slope as a vertical line. As you can see, the observed slope does not overlap even one of the 100 permuted slopes, which is another way of finding that p-value is less than 1%. This means, we reject the null hypothesis.

```{r}
ggplot(perm_slopes, aes(stat)) + geom_histogram() + # histogram distribution of slope of null hypothesis
  geom_vline(xintercept = obs_slope, color = "red") # add a vertical line for observed slope
```

### A Not-Meaningful Relationship

Now, using **mtcars** data set, let's look at the relationship between between real axle ratio (`drat`) and 1/4 mile time (`qsec`). It seems there is no relationship even without fitting a line.

```{r}
ggplot(mtcars, aes(drat, qsec)) + geom_point()
```

Let's fit a linear model anyway to estimate the linear relationship between `drat` and `qsec`. Observing the p-value, we cannot reject the null hypothesis that there is a meaningful relationship between the two variables--unlike the previous model of relationship between weight and mileage.

```{r}
tidy(lm(qsec ~ drat, data = mtcars))
obs_slope <- tidy(lm(qsec ~ drat, data = mtcars)) %>% .[2,2]
```

We can go further and visualize the null distribution of the relationship between `drat` and `qsec`. Using **infer** package, we estimate 100 permutation of the null slopes.

```{r}
perm_slopes <- mtcars %>%
  specify(qsec ~ drat) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 100, type = "permute") %>%
  calculate(stat = "slope")
```

**infer** package does a great job in simulating the null hypothesis. Next, I am going to use **ggplot2** to plot the 100 permuted slopes in a histogram, and add a red vertical line that will represent the observed slope of `qsec` and `drat`. You can see that the observed slope overlaps the permuted slopes (null hypothesis), and once again we come to realization that we cannot observe a meaningful relationship between `qsec` and `drat` variables in the **mtcars** data set.

```{r}
ggplot(perm_slopes, aes(stat)) + geom_histogram() +
  geom_vline(xintercept = obs_slope, color = "red")
```

