---
title: Plot Missing Values
author: Masood Sadat
date: '2018-05-22'
slug: plot-missing-values
categories:
  - R tips
tags:
  - R tips
---


> In the 2nd day of my **#100DaysOfCode** challenge, I want to try **visdat** package to visualize missing values. Then, I will show you how we can do a similar plot using **ggplot2** package. Then, using the second method, I will show you in addition to missing values we can visualize any other value(s).

There are many ways to inspect and find about missing values. Let's use `summary()` functions from base R to explore missing values in **airquality** dataset.

```{r message=FALSE, warning=FALSE}
summary(airquality) # summary statistics of variables/columns of airquality dataset
```

From the `summary()` function, we can see columns/variables **Ozone** has 36 missing values (NA's) and **Solar.R** has 7 missing values. We can use `View()` function to inspect the whole dataset in a spreadsheet view.

## Visualize missing values with visdat package

Let's explore the **visdat** package that promises to provide an easy way to inspect the whole dataset. If you haven't used it before, you have to install it and then load it.

```{r}
# install.packages("visdat")
library(visdat) # load visdat package
```

There are two functions in **visdat** package that we need to know: `vis_miss()` function that plots missing values of a data and `vis_dat()` function that plots class of data. After loading the **visdat** package, both functions are available and the way to use them is with a data.frame object. 

Let's use **airquality** dataset with `vis_miss()` function to visualize missing values. This plots the columns in **airquality** data.frame vertically and observations horizontally, resembling the way data is stored. Black color represents missing values and grey color represents non-missing values. There is even percentages of missing vlaues with one decimal points for each column and an overall percentage, which I find extremely useful. Compared to `summary()`, this plot gives a lot more information.

```{r message=FALSE, warning=FALSE}
vis_miss(airquality)
```

Let's use `vis_dat()` function. This plot visualizes the class of each column of **airquality** dataset in addition to indicating missing values. However, percentages of missing values are not shown. 

```{r message=FALSE, warning=FALSE}
vis_dat(airquality)
```

## Visualize missing values with ggplot2 package

Next, let's create a similar plot using **ggplot2** package. I suspect `vis_miss()` and `vis_dat()` use **ggplot2** under the hood.

I am going to load **tidyverse** package, which includes packages such as **ggplot2**, **dplyr**, and **tidyr**. So instead of loading these three packages that we need, I am going to load only **tidyverse**.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Again I am going to use **airquality** dataset. The following chunk of code produces a plot that indicates missing values. I have left comments inside the code that explains each part of the process.

```{r message=FALSE, warning=FALSE}
airquality %>%
  
  # takes airquality data.frame as argument and returns a matrix of TRUE and FALSE, with TRUE indicating missing values
  is.na() %>% 
  
  # takes the matrix as argument and converts it into a data.frame object
  data.frame() %>%
  
  # takes the data.frame object as argument, and creates a column named obs that takes value 1 through 153 which is airquality's number of rows
  mutate(obs = 1:nrow(.)) %>%
  
  # takes the data.frame object with new column obs and reshape it from wide format to long format, transforming all other columns into var and is_missing columns, with exception of obs column
  gather(var, is_missing, -obs) %>%
  
  # takes the data.frame in long format and produces a tile plot with var column in x axis, obs column in y axis, and fill the tiles based with is_missing column values
  ggplot(aes(x = var, y = obs, fill = is_missing)) + geom_tile() + 
  
  # use a black and white theme for the plot
  theme_bw() +
  
  # change x and y axes titles of the plot
  labs(x="Variables", y="Observations") +
  
  # move the x axis of the plot to the top
  scale_x_discrete(position = "top") +
  
  # change the angle of x axis labels and horizontal adjustment
  theme(axis.text.x = element_text(angle = 90, hjust = 0))
```

The above code chunk is an example of a nested code that is simplied using pipe operators **%>%** from **magrittr** package which is made avaible by loading **tidyverse** package.

## Visualize any other values with ggplot2 package

What the **ggplot2** method allows is the possibility of manipulating the code to produce plots that visualizes presence of another value. The second line of the code, `is.na()` function, transforms the **airquality** dataset into a matrix of TRUE and FALSE (or 0 and 1) which later is supplied into the `ggplot()` function to create the plot. 

Let's explore `is.na()` function separately. As you can observe, the result of `is.na(airquality)` is a matrix containing TRUE and FALSE.

```{r}
is.na(airquality) %>% 
  head() # print only first rows of the matrix produced by is.na(airquality)
```

So, instead of `is.na()` if we use another logical test (e.g. airquality > 50) to create a matrix of TRUE and FALSE too. Let's assume we are interested to know if airquality has any values greater than 50. So we bring changes in the second line `{. > 50}` and create `is_greater_than_50` instead of `is_missing` in the fifth and sixth lines. The resulting plot indicates missing values in grey color, >50 vaules with blue color, and the rest in red color. `ggplot()` has a lot of functions that allows you to customize this plot in almost any way in case you are interested.

```{r message=FALSE, warning=FALSE}
airquality %>%
  {. > 50} %>%
  data.frame() %>%
  mutate(obs = 1:nrow(.)) %>%
  gather(var, is_greater_than_50, -obs) %>%
  ggplot(aes(x = var, y = obs, fill = is_greater_than_50)) + geom_tile() + 
  theme_bw() +
  labs(x="Variables", y="Observations") +
  scale_x_discrete(position = "top") +
  theme(axis.text.x = element_text(angle = 90, hjust = 0))
```

## Create a generic function to plot any condition

Perhaps you would like to visualize every dataset you work with. In that case, you might be better off creating a function instead of typing all that code or even copy/pasting.

Let's take the existing code and turn it into a function. The following chunk of code creates a function called `missing_tiles()` that takes a data.frame and returns a tile plot of missing values.

```{r message=FALSE, warning=FALSE}
missing_tiles <- function(x){
  is.na(x) %>% 
    data.frame() %>% 
    mutate(obs = 1:nrow(.)) %>% 
    gather(var, is_miss, -obs) %>%
    ggplot(aes(x = var, y = obs, fill = is_miss)) + geom_tile() + 
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 0)) + 
    labs(x = "Variables", y = "Observations") +
    scale_x_discrete(position = "top")
}
missing_tiles(airquality)
```

Next, we create a more general function that takes any condition.

```{r message=FALSE, warning=FALSE}
```









